Let's talk about Thurstone \& Co.: An information-theoretical model for comparative judgments, and its statistical translation

### 2024-12-13 _tex/index.tex
Built site for gh-pages

In its most general form, Thurstone's theory addresses pairwise
comparisons of stimuli made by a single judge
\citep[pp.~267]{Thurstone_1927b}. The theory posits that two key factors
determine the dichotomous outcome of these comparisons: the discriminal
process of each stimulus and their discriminal difference. The
\emph{discriminal process} represents the psychological impact each
stimulus has on judges or, more simply, their perception of the stimulus
trait. According to the theory, the discriminal process for each
stimulus follows a Normal distribution, where its mode (mean), called
the \emph{modal discriminal process}, indicates the stimulus position on
the trait continuum, while its dispersion, known as the
\emph{discriminal dispersion}, reflects variability in the perceived
trait of the stimulus.

Figure~\ref{fig-discriminal_process} illustrates the discriminal
processes along a quality trait continuum for two written texts. The
figure shows that each discriminal process follows a Normal
distribution. It also indicates that the modal discriminal process for
Text B is positioned further along the continuum than that of Text A
\((S_{B} > S_{A})\), signifying a higher quality for Text B.
Additionally, the figure highlights that Text B has a broader
distribution compared to Text A, which arises from its larger
discriminal dispersion \((\sigma_{B} > \sigma_{A})\).

@@

\includegraphics[width=0.7\linewidth,height=\textheight,keepaspectratio]{./images/figures/discriminal_process.png}

@@

However, since the discriminal processes of the stimuli are not directly
observable, the theory introduces the \emph{law of comparative
judgment}. This law states that in pairwise comparisons, a judge
perceives the stimulus positioned further along the trait continuum as
having a higher level of that trait. This law emphasizes that the
relative distance between stimuli, rather than their absolute positions
on the continuum, determines the outcome of the pairwise comparison.
Indeed, the theory assumes that the observed dichotomous outcome arises
from the difference between the underlying discriminal processes of the
stimuli, referred to as the \emph{discriminal difference}. Since the
individual discriminal processes follow a Normal distribution, the
discriminal difference also follows a Normal distribution
\citep{Andrich_1978}. In this distribution, the mode (mean) represents
the relative separation between the stimuli and the dispersion captures
the variability of that relative separation.

Figure~\ref{fig-discriminal_difference} illustrates the distribution of
the discriminal difference for the texts presented in
Figure~\ref{fig-discriminal_process}. The figure indicates that the
judge perceives Text B as having significantly higher quality than Text
A. This conclusion is evident from the positive difference in their
modal discriminal processes \((S_{B} - S_{A} > 0)\) and the area under
the curve where the discriminal difference distinctly favors Text B over
Text A, denoted as \(P(B > A)\) (shaded gray area). As a result, the
dichotomous outcome of this comparison is more likely to favor Text B
over Text A.

@@

\includegraphics[width=0.7\linewidth,height=\textheight,keepaspectratio]{./images/figures/discriminal_difference.png}

@@

Since the general form of Thurstone's theory, outlined in
Section~\ref{sec-thurstone_theory}, applies to a CJ design where a
single judge evaluates multiple stimuli, Thurstone developed four
additional cases for the theory's practical application, each
progressively incorporating additional simplifying assumptions. Case I
represents the general form of the theory. Case II extends this by
allowing multiple judges to make comparisons rather than restricting the
comparisons to a single judge. Case III introduces the assumption of
zero correlation between stimuli. Case IV builds on this by assuming the
stimuli have similar dispersions. Finally, Case V replaces this
assumption with the condition that the stimuli have equal discriminal
dispersions. Table~\ref{tbl-thurstone_cases} summarizes these cases and
their assumptions. For a detailed discussion of this progression, refer
to \citet{Thurstone_1927b} and \citet[pp.~248-253]{Bramley_2008}.

@@

\includegraphics[width=0.9\linewidth,height=\textheight,keepaspectratio]{./images/tables/thurstone_cases.png}

@@

Despite its reliance on the largest number of simplifying assumptions
\citetext{\citealp[pp.~253]{Bramley_2008}; \citealp[pp.~677]{Kelly_et_al_2022}},
Case V remains the most widely used case in the CJ literature. This
popularity stems mainly from its simplified statistical representation
in the Bradley-Terry-Luce (BTL) model
\citep{Bradley_et_al_1952, Luce_1959}. The BTL model mirrors the
assumptions of Case V, with one key difference: while Case V assumes a
Normal distribution for the stimuli's discriminal processes, the BTL
model uses the more mathematically tractable Logistic distribution
\citep[pp.~254]{Andrich_1978, Bramley_2008} (see
Table~\ref{tbl-thurstone_cases}). This substitution has little impact on
the model's estimation or interpretation, as the Normal and Logistic
distributions share similar statistical properties, differing only by a
scaling factor of approximately \(1.7\)
\citep[pp.~16]{vanderLinden_et_al_2017_I} (see
Figure~\ref{fig-logistic_vs_normal}).

\begin{minipage}{\linewidth}

\includegraphics[width=0.7\linewidth,height=\textheight,keepaspectratio]{./images/figures/density.png}

\subcaption{\label{fig-density}Probability density}

\end{minipage}%
\newline
\begin{minipage}{\linewidth}

\includegraphics[width=0.7\linewidth,height=\textheight,keepaspectratio]{./images/figures/cummulative.png}

\subcaption{\label{fig-cumulative}Cummulative probability}

\end{minipage}%

\caption{\label{fig-logistic_vs_normal}Probability density and
cumulative probability of the logistic and Normal distributions.
Extracted from \citet[pp.~254-255]{Bramley_2008}.}

Nevertheless, Thurstone originally developed Case V to provide a
``rather coarse scaling'' of traits \citep[pp.~269]{Thurstone_1927b},
prioritizing statistical simplicity over precision in trait measurement
\citep[pp.~677]{Kelly_et_al_2022}. Thurstone explicitly warned against
its untested application, stating that its use ``should not be made
without (an) experimental test'' \citep[pp.~270]{Thurstone_1927b},
acknowledging that some assumptions could be problematic when
researchers asesss complex traits or heterogeneous stimuli
\citep[pp.~376]{Thurstone_1927a}. Consequently, given that modern CJ
applications frequently involve such traits and stimuli, two main
assumptions of Case V and, by extension, of the BTL model may not
consistently hold in theory or practice: the assumption of equal
dispersion and zero correlation between stimuli.

@@

The discriminal dispersions of stimuli are crucial in determining the
dichotomous outcomes of pairwise comparisons. Holding all other modeling
factors constant, discrepancies in these dispersions shape the
distribution of the discriminal difference, directly influencing the
comparison outcome. Figure~\ref{fig-dispersion} illustrates how more
uncertainty in the trait perception of one text relative to another,
\((\sigma_{B}-\sigma_{A})\), broadens the distribution of their
discriminal difference. This broadening affects the area under the curve
where the discriminal difference distinctly favors one text over the
other, \(P(B > A)\). Additionally, the figure shows that when the
discriminal dispersions of the texts are equal
\((\sigma_{B}-\sigma_{A}=0)\), the discriminal difference is more likely
to favor one text over the other than when their dispersions differ
(shaded gray area).

@@

\includegraphics[width=0.7\linewidth,height=\textheight,keepaspectratio]{./images/figures/dispersion.png}

@@

However, Thurstone contended that this assumption of equal dispersions
may not hold when researchers assess complex traits or heterogeneous
stimuli \citep[pp.~376]{Thurstone_1927a}, as these traits and stimuli
can introduce judgment discrepancies due to their unique features
\citep{vanDaal_et_al_2016, Lesterhuis_2018, Chambers_et_al_2022}.
Indeed, evidence of such violation may already exist in the CJ
literature as misfit statistics. These statistics measure the judgment
discrepancies associated with a given stimulus
\citetext{\citealp[pp.~12]{Pollitt_2004}; \citealp[pp.~20]{Goossens_et_al_2018}}.
For instance, labeling texts as ``misfits'' indicates that comparisons
involving these texts result in more judgment discrepancies than others
\citep{Pollitt_2012a, Pollitt_2012b, vanDaal_et_al_2016, Goossens_et_al_2018}.
This finding implies that discriminal differences associated with
``misfits'' texts usually display a broader dispersion, suggesting that
the discriminal processes of these texts also exhibit more variation
than other texts. Notably, this reasoning also applies to ``misfit''
judges, whose evaluations reflect substantial deviations from the shared
consensus due to the unique characteristics of the stimuli or the judges
themselves. Moreover, the presence of these ``misfit'' judges and their
deviations can introduce additional statistical and measurement issues,
which we discuss in Section~\ref{sec-theory-issue1b}.

Therefore, assuming equal dispersions between stimuli, despite its
violation, can cause Case V (and the BTL model) to inflate the
reliability of the outcome, resulting in inaccurate conclusions about
the comparison. Moreover, ignoring the disparity in dispersions can lead
to the neglect of critical differences in the reliability of the trait
across these stimuli, leading to erroneous conclusions about the trait's
estimates \citep[pp.~370]{McElreath_2020}. Additionally, if researchers
acknowledge that misfit statistics help identify these critical
differences in dispersion, the usual practice in the CJ literature of
excluding stimuli based on these statistics
\citep{Pollitt_2012b, vanDaal_et_al_2017, Goossens_et_al_2018} risks
discarding valuable information and introducing bias into the trait's
estimates \citep[chap.~12]{Zimmerman_1994, McElreath_2020}. The
direction and magnitude of these biases are often unpredictable, as they
depend on the specific stimuli researchers exclude from the analysis.

@@

Similar to the discriminal dispersions, the correlation between
discriminal processes plays a crucial role in determining the
dichotomous outcomes of pairwise comparisons. Holding all other modeling
factors constant, this correlation shapes the distribution of the
discriminal difference, directly influencing the comparison outcome.
Figure~\ref{fig-correlation} illustrates how the dependence in trait
perception between two texts narrows the distribution of their
discriminal difference. This narrowing affects the area under the curve
where the discriminal difference distinctly favors one text over the
other, \(P(B>A)\). Moreover, the figure shows that when two texts are
independent or uncorrelated \((\rho=0)\), their discriminal difference
is less likely to favor one text over the other than when the texts are
highly correlated (shaded gray area).

@@

\includegraphics[width=0.7\linewidth,height=\textheight,keepaspectratio]{./images/figures/correlation.png}

@@

Notably, Thurstone's Case V and the BTL model assume independent trait
perceptions across stimuli. Thurstone attributed this independence to
the cancellation of potential judges' biases, driven by two opposing and
equally weighted effects occurring during the pairwise comparisons
\citep[pp.~268]{Thurstone_1927b}. \citet{Andrich_1978} mathematically
demonstrated this cancellation using the BTL model under the assumption
of discriminal processes with additive biases. However, it is easy to
imagine at least two scenarios where the zero correlation assumption
almost certainly does not hold: when the pairwise comparison involves
multidimensional, complex traits with heterogeneous stimuli and when an
additional hierarchical structure is relevant to the stimuli.

@@

In any case, the psychometric and statistical literature emphasizes the
need to address factors that create dependencies between stimuli, as
failing to do so can affect the reliability of the comparison outcomes
and lead to inaccurate conclusions about the trait's estimates. For
instance, researchers who overlook additional relevant traits, such as
judges' biases, can cause dimensional mismatches in the statistical
model used for analysis. This mismatch can artificially inflate the
reliability of the trait \citep[pp.~341]{Hoyle_et_al_2023} or, even
worse, introduce bias into the trait's estimates \citep{Ackerman_1989}.
Furthermore, as discussed in Section~\ref{sec-theory-issue1a},
researchers who exclude judges based on misfit statistics can risk
discarding valuable information, further biasing the trait's estimates
\citep[chap.~12]{Zimmerman_1994, McElreath_2020}. Lastly, researchers
who fail to account for hierarchical (grouping) structures can reduce
the precision of model parameter estimates, which may amplify the
overestimation of the trait's reliability
\citep[pp.~482]{Hoyle_et_al_2023}.

@@

In aggregate, researchers' inadequate handling of violations to the
assumptions of equal dispersion and zero correlation between stimuli,
along with the apparent disconnect between CJ's approach to trait
measurement and hypothesis testing, can undermine the reliability of the
trait and ultimately compromise its validity
\citep[pp.~2]{Perron_et_al_2015}. Consequently, adopting a more
systematic and integrated approach to examining what happens when judges
compare two stimuli could offer several statistical and measurement
benefits, including addressing these issues.

@@

>>>>

In its most general form, Thurstone's theory addresses pairwise
comparisons where a single judge evaluates multiple stimuli
\citep[pp.~267]{Thurstone_1927b}. The theory posits that two key factors
determine the dichotomous outcome of these comparisons: the discriminal
process of each stimulus and their discriminal difference. The
\emph{discriminal process} captures the psychological impact each
stimulus exerts on the judge or, more simply, his perception of the
stimulus trait. The theory assumes that the discriminal process for any
given stimulus forms a Normal distribution along the trait continuum
\citep[pp.~266]{Thurstone_1927b}. The mode (mean) of this distribution,
known as the \emph{modal discriminal process}, indicates the stimulus
position on this continuum, while its dispersion, referred to as the
\emph{discriminal dispersion}, reflects variability in the perceived
trait of the stimulus.

Figure~\ref{fig-discriminal_process} illustrates hypothetical
discriminal processes along a quality trait continuum for two written
texts. The figure indicates that the modal discriminal process for Text
B is positioned further along the continuum than that of Text A
\((S_{B} > S_{A})\), suggesting that Text B exhibits higher quality.
Additionally, the figure highlights that Text B has a broader
distribution compared to Text A, which arises from its larger
discriminal dispersion \((\sigma_{B} > \sigma_{A})\).

@@

\includegraphics[width=0.7\linewidth,height=\textheight,keepaspectratio]{./images/png/discriminal_process.png}

@@

However, since the individual discriminal processes of the stimuli are
not directly observable, the theory introduces \emph{the law of
comparative judgment}. This law posits that in pairwise comparisons, a
judge perceives the stimulus with a discriminal process positioned
further along the trait continuum as possessing more of the trait
\citep[pp.~251]{Bramley_2008}. This suggests that the relative distance
between stimuli, rather than their absolute positions on the continuum,
likely defines the outcome of pairwise comparisons. Indeed, the theory
assumes that the difference between the underlying discriminal processes
of the stimuli, referred to as \emph{the discriminal difference},
determines the observed dichotomous outcome. Moreover, the theory
assumes that because the individual discriminal processes form a Normal
distribution on the continuum, the discriminal difference will also
conform to a Normal distribution \citep{Andrich_1978}. In this
distribution, the mode (mean) represents the relative separation between
the stimuli, and its dispersion indicates the variability of that
separation.

Figure~\ref{fig-discriminal_difference} illustrates the distribution of
the discriminal difference for the hypothetical texts depicted in
Figure~\ref{fig-discriminal_process}. The figure indicates that the
judge perceives Text B as having significantly higher quality than Text
A. This conclusion rests on two key observations: the positive
difference between their modal discriminal processes
\((S_{B} - S_{A} > 0)\) and the probability area where the discriminal
difference distinctly favors Text B over Text A, represented by the
shaded gray area denoted as \(P(B > A)\). As a result, the dichotomous
outcome of this comparison is more likely to favor Text B over Text A.

@@

\includegraphics[width=0.7\linewidth,height=\textheight,keepaspectratio]{./images/png/discriminal_difference.png}

@@

Thurstone observed that the general form of the theory, outlined in
Section~\ref{sec-thurstone_theory}, created a trait scaling problem. The
model required estimating more ``unknown'' parameters than the available
pairwise comparisons \citep[pp.~267]{Thurstone_1927b}. To address this
issue and facilitate the practical application of the theory, he
developed five cases derived from this general form. Each case
progressively incorporated additional simplifying assumptions into the
model.

In Case I, Thurstone assumed that pairs of stimuli maintained a constant
correlation across all comparisons. In Case II, he allowed multiple
judges to make comparisons instead of restricting evaluations to a
single judge. In Case III, he introduced the assumption of zero
correlation between stimuli. In Case IV, he assumed stimuli exhibited
similar dispersions. Finally, in Case V, he replaced this assumption
with the condition that stimuli had equal discriminal dispersions.
Table~\ref{tbl-thurstone_cases} summarizes the assumptions of the
general form and the five cases. For an in-depth discussion of these
cases and their progression, refer to \citet{Thurstone_1927b} and
\citet[pp.~248--253]{Bramley_2008}.

@@

\includegraphics[width=1\linewidth,height=\textheight,keepaspectratio]{./images/png/thurstone_cases.png}

@@

Despite relying on the most extensive set of simplifying assumptions
\citetext{\citealp[pp.~253]{Bramley_2008}; \citealp[pp.~677]{Kelly_et_al_2022}},
Case V remains the most widely used case in the CJ literature. This
popularity stems mainly from its simplified statistical representation
in the Bradley-Terry-Luce (BTL) model
\citep{Bradley_et_al_1952, Luce_1959}. The BTL model mirrors the
assumptions of Case V, with one key difference: while Case V assumes a
Normal distribution for the stimuli's discriminal processes, the BTL
model uses the more mathematically tractable Logistic distribution
\citep[pp.~254]{Andrich_1978, Bramley_2008} (see
Table~\ref{tbl-thurstone_cases}). This substitution has little impact on
the model's estimation or interpretation, as the Normal and Logistic
distributions share similar statistical properties, differing only by a
scaling factor of approximately \(1.7\)
\citep[pp.~16]{vanderLinden_et_al_2017_I}.

However, Thurstone originally developed Case V to provide a ``rather
coarse scaling'' of traits \citep[pp.~269]{Thurstone_1927b},
prioritizing statistical simplicity over precision in trait measurement
\citep[pp.~677]{Kelly_et_al_2022}. He explicitly warned against its
untested application, stating that its use ``should not be made without
(an) experimental test'' \citep[pp.~270]{Thurstone_1927b}, acknowledging
that some assumptions could prove problematic when researchers asesss
complex traits or heterogeneous stimuli
\citep[pp.~376]{Thurstone_1927a}. Consequently, given that modern CJ
applications frequently involve such traits and stimuli, two main
assumptions of Case V and, by extension, of the BTL model may not
consistently hold in theory or practice: the assumption of equal
dispersion and zero correlation between stimuli.

@@

According to the theory, the discriminal dispersions of stimuli play a
critical role in determining the outcome of pairwise comparisons.
Specifically, discrepancies in these dispersions shape the distribution
of the discriminal difference, directly influencing the comparison
outcome. Figure~\ref{fig-dispersion} illustrates this idea, assuming a
researcher can observe the discriminal processes for the texts shown in
Figure~\ref{fig-discriminal_process}. The figure also considers that the
discriminal dispersion for Text A remains constant and that the texts
have no correlation \((\rho=0)\).

Figure~\ref{fig-dispersion} reveals that more uncertainty in the trait
perception of Text B compared to Text A, \((\sigma_{B}-\sigma_{A})\),
broadens the distribution of their discriminal difference. This
broadening affects the probability area where the discriminal difference
distinctly favors Text B over Text A, expressed as \(P(B > A)\),
ultimately influencing the comparison outcome. Additionally, the figure
reveals that when the discriminal dispersions of the texts are equal
\((\sigma_{B}-\sigma_{A}=0)\), the discriminal difference is more likely
to favor Text B over Text A (shaded gray area), compared to situations
where their dispersions differ.

@@

\includegraphics[width=0.7\linewidth,height=\textheight,keepaspectratio]{./images/png/dispersion.png}

@@

In experimental practice, however, this process occurs in reverse.
Researchers first observe the comparison outcome and then use the BTL
model to infer the discriminal difference between the stimuli and their
discriminal processes \citep[pp.~373]{Thurstone_1927a}. For example,
when researchers observe a large sample of outcomes favoring Text B over
Text A and correctly assume equal dispersions between the texts, the BTL
model estimates a discriminal difference distribution that accurately
represents the ``true'' discriminal difference of the texts. This
scenario is illustrated with Figure~\ref{fig-dispersion} when the
discriminal difference distribution of the model aligns with the
``true'' distribution, represented by the thick continuous line
corresponding to \(\sigma_{B}-\sigma_{A}=0\). The estimation accuracy of
this discriminal difference, in turn, ensures reliable discriminal
process estimates for the texts (citation needed?). It then intuitively
follows that the outcome's ability to represent the ``true'' differences
between stimuli largely depends on the validity of the model's
assumptions, particularly the assumption of equal dispersions.

However, Thurstone contended that the assumption of equal dispersions
may not hold when researchers assess complex traits or heterogeneous
stimuli \citep[pp.~376]{Thurstone_1927a}, as these traits and stimuli
can introduce judgment discrepancies due to their unique features
\citep{vanDaal_et_al_2016, Lesterhuis_2018, Chambers_et_al_2022}.
Indeed, evidence of this violation may already exist in the CJ
literature as misfit statistics, which measure judgment discrepancies
associated with specific stimuli
\citetext{\citealp[pp.~12]{Pollitt_2004}; \citealp[pp.~20]{Goossens_et_al_2018}}.
For example, labeling texts as ``misfits'' indicates that comparisons
involving these texts result in more judgment discrepancies than those
involving other texts
\citep{Pollitt_2012a, Pollitt_2012b, vanDaal_et_al_2016, Goossens_et_al_2018}.
These discrepancies, in turn, suggest that the discriminal differences
for ``misfit'' texts have broader distributions, indicating that their
discriminal processes may also exhibit more variation than that of other
texts. A similar reasoning applies to ``misfit'' judges, whose
evaluations deviate substantially from the shared consensus due to the
unique characteristics of the stimuli or the judges themselves.
Moreover, these ``misfit'' judges and their deviations can introduce
additional statistical and measurement issues, which we discuss in
Section~\ref{sec-theory-issue1b}.

Then, incorrectly assuming equal dispersions between stimuli can lead
the BTL model to introduce various statistical and measurement issues.
For example, the model could overestimate the accuracy of the outcome in
reflecting the ``true'' discriminal differences between stimuli. This
overestimation may result in spurious inferences about these differences
\citep[pp.~370]{McElreath_2020} and, by extension, about the stimuli's
discriminal processes. Figure~\ref{fig-dispersion} also illustrates this
scenario when the model's discriminal difference distribution aligns
with the thick continuous line for \(\sigma_{B}-\sigma_{A}=0\), while
the ``true'' discriminal difference follows any discontinuous line where
\(\sigma_{B}-\sigma_{A} \neq 0\). Moreover, if researchers recognize
that misfit statistics highlight critical differences in dispersions,
the common practice in CJ literature of excluding stimuli based on these
statistics
\citep{Pollitt_2012a, Pollitt_2012b, vanDaal_et_al_2016, Goossens_et_al_2018}
may inadvertently discard valuable information, introducing bias into
trait estimates \citep[chap.~12]{Zimmerman_1994, McElreath_2020}. These
biases are often unpredictable, as they depend on the specific stimuli
excluded from the analysis.

@@

Denoted by \(\rho\), the correlation measures the dependence of a
judge's perception of the trait in one stimulus on his perception of the
same trait in another. Like the discriminal dispersions, this
correlation shapes the distribution of the discriminal difference and
directly influences the outcomes of pairwise comparisons.
Figure~\ref{fig-correlation} illustrates this concept, assuming the
researcher can observe the discriminal processes for the texts shown in
Figure~\ref{fig-discriminal_process}. The figure also considers that the
discriminal dispersions for both texts remain constant.

Figure~\ref{fig-correlation} reveals that as the correlation between the
texts increases, the distribution of their discriminal difference
becomes narrower. This narrowing affects the area under the curve where
the discriminal difference distinctly favors Text B over Text A, denoted
as \(P(B > A)\), thus influencing the comparison outcome. Furthermore,
the figure shows that when two texts are independent or uncorrelated
\((\rho=0)\), their discriminal difference is less likely to favor Text
B over Text A (shaded gray area), compared to scenarios when the texts
are highly correlated.

@@

\includegraphics[width=0.7\linewidth,height=\textheight,keepaspectratio]{./images/png/correlation.png}

@@

However, in experimental practice, researchers typically follow this
process in reverse. For example, when they observe a large sample of
outcomes favoring Text B over Text A and correctly assume zero
correlation between the texts, the BTL model estimates a discriminal
difference distribution that accurately represents the ``true''
discriminal difference of the texts. This scenario is illustrated with
Figure~\ref{fig-correlation} when the discriminal difference
distribution of the model aligns with the ``true'' distribution,
represented by the thick continuous line corresponding to \(\rho=0\).
The accuracy of this discriminal difference estimation, in turn, ensures
reliable estimates for the discriminal process of the texts (citation
needed?).

Notably, Thurstone's Case V and the BTL model assume independent
discriminal processes across comparisons. Thurstone attributed this
independence to the cancellation of potential judges' biases, driven by
two opposing and equally weighted effects occurring during the pairwise
comparisons \citep[pp.~268]{Thurstone_1927b}. \citet{Andrich_1978}
mathematically demonstrated this cancellation using the BTL model under
the assumption of discriminal processes with additive biases. However,
it is easy to imagine at least two scenarios where the zero correlation
assumption almost certainly does not hold: when the pairwise comparison
involves multidimensional, complex traits with heterogeneous stimuli and
when an additional hierarchical structure is relevant to the stimuli.

@@

{In any case, similar to Section~\ref{sec-theory-issue1a}, incorrectly
assuming zero correlation between stimuli can lead the BTL model to
introduce various statistical and measurement issues. For instance, the
model could over- or underestimate the accuracy of the outcome in
reflecting the ``true'' discriminal differences between stimuli. This
over- underestimation may result in spurious inferences about these
differences and, by extension, about the stimuli's discriminal processes
\citep[pp.~341]{Hoyle_et_al_2023}. Figure~\ref{fig-correlation} also
illustrates this scenario when the model's discriminal difference
distribution aligns with the thick continuous line for \(\rho=0\), while
the ``true'' discriminal difference follows any discontinuous line where
\(\rho \neq 0\). This missaligment can be due to the overlook of
additional relevant traits, such as judges' biases, which cause
dimensional mismatches in the BTL model, artificially inflating the
reliability of the trait \citep[pp.~341]{Hoyle_et_al_2023} or, even
worse, introduce bias into the trait's estimates \citep{Ackerman_1989}.
Furthermore, researchers who exclude judges based on misfit statistics
can risk discarding valuable information, further biasing the trait's
estimates \citep[chap.~12]{Zimmerman_1994, McElreath_2020}. Lastly,
researchers who fail to account for hierarchical (grouping) structures
can reduce the precision of model parameter estimates, which may amplify
the overestimation of the trait's reliability
\citep[pp.~482]{Hoyle_et_al_2023}.}

@@

{In aggregate, researchers' inadequate handling of violations to the
assumptions of equal dispersion and zero correlation between stimuli,
along with the apparent disconnect between CJ's approach to trait
measurement and hypothesis testing, can undermine the reliability of the
trait estimates and ultimately compromise its validity
\citep[pp.~2]{Perron_et_al_2015}. Consequently, adopting a more
systematic and integrated approach to examining what happens when judges
compare two stimuli could offer several statistical and measurement
benefits, including addressing these issues.}

@@

The (latent) discriminal difference of the stimuli directly determines
the (manifest) outcome of the pairwise comparisons

\includegraphics[width=0.29\linewidth,height=\textheight,keepaspectratio]{./images/png/CJ_TM_A1.png}

\caption{\label{fig-CJ_TM_A1}Theoretical model A1\$}

The (latent) ``perceived'' discriminal processes for the stimuli
directly determines their discriminal difference

\includegraphics[width=0.4\linewidth,height=\textheight,keepaspectratio]{./images/png/CJ_TM_A2.png}

\caption{\label{fig-CJ_TM_A2}Theoretical model A2\$}

The (latent) ``true'' discriminal processes for the stimuli and the
judges' biases directly determines their (latent) ``perceived''
discriminal processes

\includegraphics[width=0.6\linewidth,height=\textheight,keepaspectratio]{./images/png/CJ_TM_A3.png}

\caption{\label{fig-CJ_TM_A3}Theoretical model A3\$}

without loosing generality, the (latent) ``perceived'' and ``true''
discriminal processes for the stimuli can be depicted in a vector for
each judge, as in
